The ChEMBL database as Linked Open Data

### 2013-01-15 article.tex
fix minor grammar

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped the ChEMBL dataset to the Resource Description Framework (RDF), resulting
in Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples, reflecting ``facts''. A fact can
be given on any topic, and RDF therefore talks about \textit{things} very generically~\cite{Miller:04:RP}. Things in RDF
are represented by Universal Resource Identifiers (URI). When the URI can be resolved, 
for example using the HTTP protocol, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not all, domains, making it a successful 
approach to data integration in the life sciences.

This paper presents an update on the ChEMBL-RDF resource, including details of the latest structures
and ontologies used to map ChEMBL to RDF, along with new use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

@@

It should also be noted that all drug-like entities in ChEMBL are not ontologically
subclassing a drug class. Instead, we give the entities the role of drug: only within a
particular context, they would be drug. If not used as medication, they are not drug.
Moreover, the triples only attach this drug role to approved drugs, as defined by the
ChEMBL database. All drug-like entities in the database do not have the role of drug.
The role is triplified using the OBO and ChEBI ontologies in the following manner,
where CHEBI\_23888 is the ontological entry for of ``drug role'' in ChEBI:

@@

Each release of the ChEMBL database is accompanied by a set of release counts, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

The following SPARQL query returns 1,143,682 entities for which activities are captured in ChEMBL:

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p bfo:MaterialEntity .
}
\end{verbatim}
\end{small}

The following SPARQL query returns 617,681 ChEMBL assays:

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Assay .
}
\end{verbatim}
\end{small}

It should be noted that this returns five entries less than the SQL database, because the
triple generation requires the assay type to be defined, which is not the case for those 
five assays.

The following SPARQL query returns the 6,933,068 ChEMBL activities:

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p chembl:Activity .
}
\end{verbatim}
\end{small}

The following SPARQL query returns the 44,681 ChEMBL documents. This is one fewer than appears 
in the ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as the result does not include the doc\_id -1, as this refers to the 
unpublished dataset, not itself a valid \verb+bibo:Article+.

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p bibo:Article .
}
\end{verbatim}
\end{small}

@@

ChEMBL-RDF can be used to explore interrelated scientific datasets, and we here present a few applications. 
The first application describes how the SPARQL end point can be used to make the data available as linked data
via the Bio2RDF platform. The second example shows how Chem2Bio2RDF can combine the ChEMBL data
with other life sciences databases, showing the power of the Linked Data approach.
The third application takes advantage of the bibliographic information exposed as machine readable
data, in calculating citation statistics. The fourth use case shows an integration of ChEMBL-RDF with
ChemSpider to provide an extension for the decision support platform in Bioclipse. The final application
shows a more complex SPARQL query showing how information from various tables can be used to
find compounds with a high selectivity.

@@

A second resource in the linked open data network into which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF is a single RDF repository covering over twenty public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative searches~\cite{Chen2012a} and
data mining~\cite{Chen2012b}. The ChEMBL-RDF set has been uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL-RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link the two. For instance,
to investigate the relations between drug side effects and their targets (usually off-target effects), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart disease were linked to 87 unique protein targets ($IC_{50}$ $<$ 10 $\mu$m). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and SIDER in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

@@

Using the SPARQL end point, it is easy to integrate data from ChEMBL into other web resources, such as CitedIn (http://www.citedin.org)~\cite{CitedIn2011}.
While the impact of a scientific publication is traditionally measured by frequency of citation in subsequent publications, this
narrowly defined calculation increasingly fails to capture the true reach of a paper. CitedIn was developed to make scientific
impact more visible and more accurate by providing insight into where published work is being cited online in a variety of public
resources. ChEMBL contains many literature references. The SPARQL endpoint is queried by CitedIn and the results are presented to its users.
Figure~\ref{fig:citedin} shows the results for the Nature paper by Gamo presenting the malaria structures~\cite{Gamo2010}
which is cited as the reference for a large number of activity datapoints and compounds in ChEMBL.
Counting all objects that reference this paper is done by the following SPARQL query:

@@

Two ontologies were recently developed that may address these issues for ChEMBL-RDF in future. Firstly, the BioAssay Ontology, which addresses many of our
remaining concerns around making the experimental data machine readable~\cite{Visser2011}. Secondly,
the Quantity, Unit, Dimension and Type (QUDT) ontology which makes it possible
to express measurements in a standardised machine readable way~\cite{QUDT}. Both approaches are 
being explored in the Open PHACTS project and will be applied to ChEMBL-RDF as the Open PHACTS project progresses~\cite{Williams2012}.

@@

The ongoing ChEMBL-RDF project is Open Source making it possible to others to convert ChEMBL to RDF too.
The ChEMBL team (JO, AG, MD) plans to be a primary source of a RDF version of ChEMBL, while others continue
exposing ChEMBL as RDF too (EW, PA). Various version may arise, exposing the data in different ontologies,
supporting community needs, though convergence may arise after a period of experimenting with ontologies that
will not survive the test of time. The availability of Open Source scripts ensure an easy switch from one
ontology to another.

>>>>

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped the ChEMBL dataset to the Resource Description Framework (RDF), resulting
in Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples, reflecting ``facts''. A fact can
be given on any topic, and RDF therefore talks about \textit{things} very generically~\cite{Miller:04:RP}. Things in RDF
are represented by resources named with Universal Resource Identifiers (URI). When the URI can be resolved, 
for example using the HTTP protocol, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not all, domains, making it a successful 
approach to data integration in the life sciences.

This paper presents an update on the ChEMBL-RDF data set, including details of the latest structures
and ontologies used to map ChEMBL to RDF, along with new use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

@@

It should also be noted that all drug-like entities in ChEMBL are not ontologically
subclassing a drug class. Instead, we give the entities the role of drug: only within a
particular context, they would be drug. If not used as medication, they are not drug.
Moreover, the triples only attach this drug role to approved drugs, as defined by the
ChEMBL database. All drug-like entities in the database do not have the role of drug.
The role is triplified using the OBO and ChEBI ontologies in the following manner,
where CHEBI\_23888 is the ontological entry for ``drug role'' in ChEBI:

@@

Each release of the ChEMBL database is accompanied by a set of release statistics, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

A similar SPARQL query, replacing \verb+chembl:Target+ with \verb+bfo:MaterialEntity+, returns 1,143,682 entities for which activities are captured in ChEMBL, and by using \verb+chembl:Assay+, the query returns the total of 617,681 ChEMBL assays.
It should be noted that last query returns five entries less than the SQL database, because the
triple generation requires the assay type to be defined, which is not the case for those 
five assays. With \verb+chembl:Activity+ and \verb+bibo:Article+, we count 6,933,068 ChEMBL activities and 44,681 documents,
respectively. The document count is one fewer than appears 
in the ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as the result does not include the doc\_id -1, as this refers to the 
unpublished dataset, not itself a valid \verb+bibo:Article+.

@@

ChEMBL-RDF can be used to explore interrelated scientific datasets, and we here present five applications,
describing how the SPARQL end point can be used, the combination of the ChEMBL data
with other life sciences databases, the use of the bibliographic information in calculating citation
statistics, and how to integrate ChEMBL-RDF with ChemSpider to provide an extension for the decision
support platform in Bioclipse.

@@

A second resource in the linked open data network into which ChEMBL-RDF was integrated is Chem2Bio2RDF.
Chem2Bio2RDF is a single RDF repository covering over twenty public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative searches~\cite{Chen2012a} and
data mining~\cite{Chen2012b}. The ChEMBL-RDF set was uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL-RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link the two. For instance,
to investigate the relations between drug side effects and their targets (usually off-target effects), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart disease were linked to 87 unique protein targets ($IC_{50}$ $<$ 10 $\mu$m). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can also be linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and SIDER in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

@@

Using the SPARQL end point, it is easy to integrate data from ChEMBL into other web resources, such as CitedIn (http://www.citedin.org)~\cite{CitedIn2011}.
While the impact of a scientific publication is traditionally measured by frequency of citation in subsequent publications, this
narrowly defined calculation increasingly fails to capture the true reach of a paper. CitedIn was developed to make scientific
impact more visible and more accurate by providing insight into where published work is being cited online in a variety of public
resources, as listed on the CitedIn website~\cite{CitedResources}.
ChEMBL contains many literature references. The SPARQL endpoint is queried by CitedIn and the results are presented to its users.
Figure~\ref{fig:citedin} shows the results for the Nature paper by Gamo presenting the malaria structures~\cite{Gamo2010}
which is cited as the reference for a large number of activity datapoints and compounds in ChEMBL.
Counting all objects that reference this paper is done by the following SPARQL query:

@@

Two ontologies were recently developed that may address these issues for ChEMBL-RDF in future.
Firstly, the BioAssay Ontology, which addresses many of our
remaining concerns around making the experimental data machine readable~\cite{Visser2011}. For example,
it provides an ontology of activity types. Secondly,
the Quantity, Unit, Dimension and Type (QUDT) ontology which makes it possible
to express measurements in a standardised machine readable way~\cite{QUDT}. Both approaches are 
being explored in the Open PHACTS project and will be applied to ChEMBL-RDF as the Open PHACTS project progresses~\cite{Williams2012}.

@@

The ongoing ChEMBL-RDF project is Open Source making it possible to others to convert ChEMBL to RDF too.
The ChEMBL team (JO, AG, MD) plans to be a primary source of a RDF version of ChEMBL, while others continue
exposing ChEMBL as RDF too (EW, PA). Various version may arise, exposing the data in different ontologies,
supporting current community needs, however, convergence is expected after a period of experimenting with ontologies that
may not prove to be practically useful. The availability of Open Source scripts ensures that the community can contribute to the
process of choosing a practically useful set of ontological terms.
